Have you forgotten?' Conversion experiences in shopping lists

### 2017-02-09 lists_conversion_experience.rmd
done 10.5 pages

Amidst the predictive conditions of contemporary media-culture, the inevitable incompleteness of lists is both a problem and a potential. 
This paper investigates practices of list-making associated with online grocery shopping at a  UK and global supermarket chain, Tesco. 
It will analyse how a recommender system suggests a small number of grocery items of personal relevance to each of the millions of online grocery shoppers. 
The predictive modelling practices aim to up-sell and cross-sell some of the 250,000 items in the supermarket's inventory. 
I find it interesting to approach the list-making practices  of the grocery recommender system as a mundane case of the predictive intermediaries operating in many places. 
The transformation or conversion of listing into predicting can be analysed in recommender systems that extend existing lists with potential new items, thus transforming the temporal and agential fabric of the list.
While all written lists mix past and future actions, actual and potential events, their predictive re-writing in recommender systems adds a new operational dynamic.
In this operational dynamic, the incompleteness of lists, their propensity to expand or change, becomes more important than their capacity to exhaustively enumerate things. 
The conversion of lists from finite to in-definite in recommender systems suggests we might need to re-conceptualise how we think about social order and structure in media cultures.

## Introduction: the presentation of a presentation

I saw recently a presentation about online grocery shopping at Tesco, a large supermarket chain based in the UK. The lists in this setting are recommendations for further grocery purchases under the title of 'Have you forgotten?'. When you shop for groceries online, a list of 'Have you forgotten' recommendations appear  at the checkout stage. While the title 'have you forgotten' implies that you will be shown items that could have been or are usually on a shopping list, the working assumption is a temporal one -- the list is incomplete either because you have forgotten to list something, or you didn't know what should be on your list, perhaps because you have never even thought of putting that thing on your list.

@@

The items on this list could be recommendations. They are only a few of them: five. The question: which five items and in what order?

![shopping_list](figure/1_shopping.png )

Shopping lists such as Figure \ref{fig:shopping_list} are quite prosaic -- a dozen lines scrawled on the back of a used envelope stuck on the refrigerator door  -- and sublime -- a million products ranked in some kind of order for tens or hundreds of millions of people. Such lists are highly repetitive, yet mix items of great diversity and display complicated temporal structures, spanning past and future, actual and potential. I regard the list of five suggestions show to Tesco's online customers at the checkout as exemplifying a major contemporary form of ordering, the recommendation list, that we see in many  forms of news, media, entertainment, marketing, advertising.

@@

A recommendation system should suggest items that you have not forgotten but may have never bought. If you do buy recommended items, then the recommender system has been effective in shunting the situation differently. It has led to a **'conversion.'** From a supermarket perspective, the conversion should be high in price. There is little value in successfully recommending a whole lot of small and cheap items such as rubber bands or toothpicks. So it might need to choose amongst different recommendations and weight high value items differently to low value items. In either case, 'conversion'  is the valued event.

## How do we approach lists as objects of research?

While lists such as recommendations intersect strongly with other meaning and sense-making practices in everyday life and popular culture, they are also somewhat refractory to discourse  and textual analysis methods. On the one hand, lists are highly associative. They can be interpreted or decoded semiotically, but they also exhibit great variations in textuality that multiply semiotic readings. On the other hand, as operational inscriptions, they can be treated ethnomethodologically, as the production of ('the order of social settings through ...  shared sense making practices.')[https://en.wikipedia.org/wiki/Ethnomethodology]

List practices such as online grocery recommendations figure as sites of intersection between people, infrastructures, things, places and predictions.  We might, therefore,  approach lists from various novel angles -- as asignifying semiotics as Maurizio Lazzarato calls it in his _Signs and Machines_ [@Lazzarato_2014] or as elements in 'a new order of spatio-temporal continuity for forms of economic, political and cultural life' as Celia Lury terms in her account of topological turn [@Lury_2012]. In this paper, I focus on lists and list-making in terms of the predictive conditions of contemporary media experience [@Hansen_2015a].

The question for me here is **how** contemporary lists can still be understood as produced through 'shared sense-making practices.' All lists are forms of order, since even random ordering is operationally important, but what counts as order might historically transformable, in which case the order of social settings or the problem of social order might be deeply historically contingent. The problem of order lies at the core of lists making, and our relation to lists can be seen as a test of the way that order operates more generally.

In this context, the fact that the lists and indeed supermarkets more generally are driven by database logics suggests that working with list data could well entail interacting with databases.

This is no mere methodological problem if, as Mark Hansen has recently suggested,

@@

As the quote shows, he presents this as an 'ontological transformation.' The practical and indeed experimental question for me here is whether such transformations can be detected and articulated in the prosaic setting of shopping lists and  shopping baskets. I will suggest that 'conversion' might be a better way of mapping these changes than 'transformation.'

## 1. The context of the presentation: 'statistics and operations research in industry'

I heard this presentation at an academic event with speakers from industry and government on statistics and operations research in industry. Speakers came from Shell, Disney, GCHQ, and various universities.  Because I've done a lot of work on machine learning and data science more generally, I found the sessions both familiar and occasionally exciting.

![patel_notes](figure/patel_overall.png)

Shreena Patel works as a data scientist for DunnHumby, a well-known [ customer science company ](https://www.dunnhumby.com/)  focusing on online grocery shopping at Tesco. I found her presentation on shopping lists to be  practical and evocative since it was focused quite a lot on implementation of recommendations, and symptomatic in various ways. It was filled with graphs, numbers, and tables concerning different aspects of the 'have you forgotten?' list. Understandably, this was not presented in a language of ontological transformation.

To this end, I will be seeking to reconstruct some of the workings of the recommender system by generating queries, and exploring results from  a very large supermarket chain database.

```{r simulate_basket, echo=FALSE, cache=TRUE, fig.margin=TRUE, fig.cap='Prices and quantities of things in the world'}
library(dplyr)
library(ggplot2)
library(readr)
library(knitr)

@@

g = ggplot(df, aes(x=price,y=ContentsQuantity)) + geom_point(alpha=0.3, size=0.9 ) + theme(axis.text=element_text(size=10))  + xlab('Price (£)')
g + scale_x_log10() + geom_smooth()
```

```{r  simlate_basket_2, echo=FALSE, cache=TRUE, fig.margin=TRUE}
s = sample_n(df[df$price>=30,], size=20)
kable(s[, c('name','price')])
```

Much of my paper, and my responses to the presentation I saw from a data scientist  will focus on how the ordering that is done in a recommendation lists relates to the problem of social order more generally.  The main narrative of the presentation concerned a shift from a set-based model for making lists to a predictive, probabilistic way of making lists in almost realtime and dealing with lists on a scale that is often difficult to grasp concretely. The change in list-making associated with learned models is increasingly widespread, and relies on a probabilistic treatment of people and things.

I respond to the presentation in two ways. First, by recounting, contextualising and commenting on some of the main steps in making the 'Have your forgotten?' list  as a example of the sense-making done by data scientists and customer analytics teams in many places. Many of these steps were mentioned only briefly in the presentation, and they are largely taken for granted as part of predictive analytic practice.  Second, by trying to step into the list-making practices  as an occasion to intervene and participate in the ontological conversion practices they engage in.

## 2. The web and the shopping basket: how it was reconstituted in the 1990s

!['Every little bit helps':Tesco ClubCard](figure/tesco-clubcard.jpg )

Tesco is the largest supermarket chain in UK, famous for Clubcard starting in 1991; often cited as a leading early example of Big Data Analytics. 
DunnHumby -- Edwina Dunn and Clive Humby -- are famous for having convinced the CEO of Tesco sometime in 1991 that a loyalty card program could change the supermarket chain's relationship to its customers. [@OKeefe_1984]  Clive Humby's academic work is hard to track down, but an early paper of his given at the Conference of Your Operational Researchers in Nottingham in 1984 suggests the direction that he and later Tesco would take in constructing lists [@OKeefe_1984]

@@

While Tesco has been very successful in data-mining its customers,  and perhaps became the UK's biggest supermarket through data-mining in the 1990s, the online shopping situation is a bit different. It is no longer organised around campaigns involving special offers or redemption of points, but on continuous up-selling and cross-selling to customers who might be anywhere, but especially are not in the supermarket itself. .

>Tesco is the clear winner in the online grocery market, in fact it takes almost 50p of every £1 spent on food shopping on the internet. [ https://econsultancy.com/blog/64841-how-seo-helps-tesco-to-dominate-the-online-grocery-market/ ]( https://econsultancy.com/blog/64841-how-seo-helps-tesco-to-dominate-the-online-grocery-market/ )

@@

How should characterise this shift from ClubCard membership data mining to online grocery markets?  Patel described her work at DunnHumby in terms of a shift in models from a 'rules-based list' to a 'relevancy model.' This was a significant change in how the 'Have You Forgotten' list was constructed from one kind of model to another. For me, this shift is a part of broader re-organisation of prediction in which existing predictive practices are being re-distributed and intensified in particular ways. The shift in models matters since it amounts to a much more probabilistic structuring of lists.

>>>>

This paper investigates practices of list-making associated with online grocery shopping at a  UK and global supermarket chain, Tesco. 
It reconstructs an account of how a recommender system suggests a small number of grocery items of personal relevance to each of the millions of online grocery shoppers. 
The recommender system  aim to up-sell and cross-sell some of the 250,000 items in the supermarket's inventory. 
I suggest that we can understand devices such as recommender systems as transformations or conversions of the temporal and agential fabric of the important social ordering device, the list.
While all written lists mix past and future actions, actual and potential events, their predictive re-writing in recommender systems adds a new operational dynamic.
The incompleteness of lists, their propensity to expand or change, becomes more important than their capacity to  enumerate or rank things. 
The conversion of lists from finite to in-definite in recommender systems suggests we might need to re-conceptualise how we think about social order and structure in media cultures.

##

Online grocery shopping at Tesco, a large supermarket chain based in the UK, includes recommendations for further grocery purchases under the title of 'Have you forgotten?'. When customers shop for groceries online, a list of five recommendations appear  at the checkout stage. These recommendations are the product of a recommender system, an important category of operational device in big data (see for instance [@Striphas_2015] for analysis of Netflix recommendation; [@Morris_2015]  or [@Seaver_2015] for an account of music recommendations]).  The question 'have you forgotten?' is followed by some  grocery items that could have been or are usually on a shopping list. The title of the suggestions is a bit misleading: the recommender system, as we will see, is not concerned with forgetting, with the many slips and oversights associated with shopping lists,  but rather with items that you didn't know what should be on your list, perhaps because you have never thought of putting them there.

The shelves of large contemporary supermarket are the endpoint of global logistic supply chains, in all their logistical, value-transforming and brand-mediated hypercomplexity (see [@Nielsen_TBA; @Tsing_2009] on logistics and power are the endpoint of global logistic supply chains, in all their logistical, value-transforming and brand-mediated hypercomplexity (see [@Nielsen_TBA; @Tsing_2009] on logistics and power)) Groceries display a geography of agriculture, industry, transport, communication and financialisation driven by flows of labour and capital. Encounters between this hyper-complex geography and people, even in the confines of a supermarket, are no simple matter, either for shoppers or for supermarket operators such as Tesco. While there are many facets to shopping [@Miller_2013], the shopping list, whether written on the back of an envelope, or saved as a list in an online grocery shopping system, is one intersectional ordering device. In a shopping trajectory (trawling along aisles packed with thousands of products, scrolling down screens or searching for particular brands amidst search results), lists inscribe some kind of order that filters or reduces the excessive propensities, claims, dazzle and distraction of commodities.

Hand-written shopping lists undoubtedly have ongoing importance and display interesting mixed ordering practices (see the list of handwritten shopping lists at [Grocery List](http://www.grocerylist.org)), online shopping lists take shape at the intersection of web and internet infrastructures, supply chain logistics, individualised practices and *habitus,* and increasingly, the predictive operations of  recommender systems. Just as the aisles and shelves of a supermarket present a commercially ordered geography of the world,  online shopping recommender systems generate lists that seek to align people to commodities they otherwise might have little relation to.

@@

The items on this list could be recommendations. They are only a few of them: five. The question: which five items and in what order?
```{r shopping-hand}
include_graphics('figure/1_shopping.png')
```

In terms of content, there are many similarities between shopping lists such as Figure \@ref(fig:shopping-hand)  -- a dozen lines scrawled on the back of a used envelope stuck on the refrigerator door  --  and the five ranked items shown to a customer on an online shopping site (as shown in Table \@ref(tab:random-basket)). While they are similar in content,  the list of five suggestions shown to Tesco's online customers at the checkout  exemplify a transformation in forms of ordering, a transformation that may affect the ontological texture of the world in the same way that commodification does. In its re-writing of the list, the recommender system effects a conversion experience in which individualised desire, consumption practices, value-chains, infrastructural time-space dynamics, the lives of others and the propensities of things all come into play.   ing a major contemporary form of ordering, the recommendation list, that we see in many  forms of news, media, entertainment, marketing, advertising.

@@

Despite being titled 'Have You Forgotten?' Tesco's recommender system does not suggest forgotten items.  It focuses on 'conversion events.' If a shopper does buy  recommended items, 'conversion' has occurred.  From a supermarket perspective, high-value conversion are desirable. There is little value in successfully recommending small or cheap items such as rubber bands or toothpicks.  From a data science perspective, conversion events, when they occur, confirm the power of predictive analytics to change the world in some way. From a sociological perspective, constructing and configuring recommender system in the name of conversion events, the recommender system effects an epistemic-operational transformation central to big data as an operational formation that responds to an ongoing control crisis [@Beniger_1986]. This central transformation is that the idea that data affords apprehension, visibility and management of associative propensities in the world.  Recommender systems, then, are concerned with the conversion of social order.

## Lists in and as data

Lists have long interested historians, anthropologists, sociologists, literary and media theorists. If, as the social anthropologist Jack Goody argued, lists are historically primary as forms of writing [@Goody_1986], and if list-making and its later variants (e.g. tables) precede discursive and narrative writing practices, then we might expect lists to function as powerful social ordering devices. As literary scholars suggest (see [@Mainberger_2003]), although lists have often been de-valued as literary forms, list-making often appears, even in literary form, whenever writing seeks to address, name, group or evoke totality, profusion, excess or abundance. From both  anthropological and literary perspectives, shopping lists have excellent reasons to exist: they are deeply rooted in urban social formations historically and culturally. Lists such as online grocery recommendations weave together people, infrastructures, things, places and predictions.[^1.456]

[^1.456]: While lists intersect strongly with other meaning and sense-making practices in everyday life and popular culture, they can be refractory to discourse  and textual analysis methods built around models of language or speech, with its rules of syntax and grammar. On the one hand, lists are highly associative. They can be interpreted or decoded semiotically, although they exhibit  variations in textuality -- how they are written and read -- that thwart semiotic readings. On the other hand, as operational inscriptions, they can be treated ethnomethodologically, as the production of social order in a given setting.

We might, in the light of their mutability, permeability and embedding in social order,   approach lists from various theoretical angles -- as asignifying semiotics as Maurizio Lazzarato calls it in his _Signs and Machines_ [@Lazzarato_2014] or as elements in 'a new order of spatio-temporal continuity for forms of economic, political and cultural life' as Celia Lury puts it in her account of topological turn [@Lury_2012]. In this paper, I focus on lists and list-making in terms of the predictive conditions of contemporary media experience [@Hansen_2015a].

All lists are forms of order. Mathematicians distinguish lists from sets on the basis of order. Formally, both lists and sets are collections, but a list also contains a mapping of its elements to an ordering, usually the natural or counting numbers. Sometimes this order is explicit. Numbers are written down one side of the list. Often the order is implicit to the spatial arrangement of the list.  Even random order can be  operationally important (for instance, in gambling machine or a psychology experiment, randomly order lists might be important).  Mappings between list order and social order are manifestly highly contingent. Listing practices in a given social setting, however, always index its engagement with necessary unstable dynamics of social order. From the perspective of social order, shopping lists, as they are co-constructed by a supermarket chain's recommender system, might be important intersectional zones for the conversion of the average everydayness of lists through data-analytic propensities.

Mark Hansen has recently suggested,

@@

Hansen predicates prediction as an 'ontological transformation' transforming the relation between probabilities and things.  The practical and indeed empirical  question is whether such transformations in probability that bring 'the propensities of things'  can be detected and articulated in the prosaic setting of shopping lists and  shopping baskets. 'Conversion' might be a preferable term for these changes.  It highlights a range of re-orientations in subjects, experience, things, numbers and structures  entailed more than the subjective 'understood-to-be-expressed' formulation that Hansen emphasises.

## DunnHumby's work for Tesco

In one of the many industry-meets-academia events occurring in increasingly data science-oriented higher education institutions in the UK, speakers from industry, government and commerce described their work.[^11.1] Shreena Patel, a PhD graduate in statistics and operations research,  works as a data scientist for DunnHumby, a well-known [ customer science company ](https://www.dunnhumby.com/)  focusing on online grocery shopping at Tesco.  Speaking to an audience of statisticians and operations researchers, Patel   focused on the models underlying  shopping list recommendations. Her presentation was filled with graphs, numbers, and tables concerning different aspects of the 'Have you forgotten?' recommender system. Patel's presentation was not sociological. In order to explore the operational environment she described, and its predictive operations, I will reconstruct key features of her presentation using sample data drawn from the Tesco API ([Application Programmer Interface](https://devportal.tescolabs.com/docs/services/56c73b1bf205fd0ed81dbe7a/operations/56d5b2bee2813e0a0053c47c), [@Tesco.com_2017])  A sample of the most highly priced items in Tesco stores is shown in Table  \@ref(tab:expensive-groceries) and a range of data science, statistics and graphics libraries implemented in the widely-used statistical programming language `R` [@RDevelopmentCoreTeam_2010].[^11.2]

[^11.1]: More than 100  Data Science Institutes have been set up in North America, Europe and UK since around 2012. The traffic between higher education and data analytics in industry is intense and flows in several forms: people funding, research findings, software and technical devices (code) and training.

[^11.2]:

```{r simulate-inventory, echo=FALSE, cache=TRUE, fig.margin=TRUE, fig.cap='Prices and quantities of things in the world'}
library(dplyr)
library(ggplot2)
library(readr)
library(knitr)

@@

g = ggplot(df, aes(x=price,y=ContentsQuantity)) + geom_point(alpha=0.3, size=0.9 ) + theme(axis.text=element_text(size=10))  + xlab('Price (£)')
g + ylab('Number of products') + scale_x_log10() + geom_smooth()
```

Figure \@ref(fig:simulate-inventory) plots something of the density of commodities in the Tesco grocery database. By virtue of their sheer abundance, the plot cannot easily separate the many items that Tesco sells, especially in the price range £1-£10.

```{r  expensive-groceries, echo=FALSE, cache=TRUE, fig.margin=TRUE}
s = sample_n(df[df$price>=30,], size=10)
kable(s[order(s$price), c('name','price')])
```

I respond to Patel's presentation in two ways. First, by recounting, contextualising and commenting on some of the main steps in making the 'Have your forgotten?' list  as a example of the predictive sense-making done by data scientists and customer analytics teams working with transactional data in many commercial settings. Patel mentioned many of these steps only fleetingly in the presentation, for they are largely taken for granted as part of predictive analytic practice.  Second, by trying to step into the list-making practices  as an occasion to apprehend in the wider conversions they signal. As a sociological response to a data scientist presenting a model to an audience of statisticians and operations researchers, my interest lies less in specific technical innovations and more on how the ordering work done by recommender systems relates to problems of social order more generally.  The method of reconstruction borrows from the philosopher John Dewey a broad interest in how practical and abstract elements of experience are interwoven [@Dewey_1957]. More concretely, it draws heavily on the vision of 'interface methods' proposed by Noortje Marres [@Marres_2015] in order to keep in view both the highly technical orderings of lists in recommender system and the changes in social orderings they might imply or impel.

## How the shopping list was reconstituted in the 1990s

The main narrative of Patel's presentation concerned a shift from a well-established demographically-targeted data mining model for making recommendations to a predictive, probabilistic, personally relevant  re-writing of the shopping list in almost-realtime. The changes Patel described are increasingly widespread. While they are configured in Tesco-specific ways by DunnHumby (and this reflects a longer history), they are also broadly symptomatic. Tesco is the largest supermarket chain in UK, famous for its customer loyalty and targeted marketing programme known as 'Tesco  Clubcard,' which started in 1991. DunnHumby -- founded by operations researchers Edwina Dunn and Clive Humby -- is famous for having convinced the CEO of Tesco sometime in 1991 that a loyalty card program could change the supermarket chain's relationship to its customers.  Clive Humby's academic publications  are hard to track down, but an early paper given at the _Conference of Young Operational Researchers_ in Nottingham in 1984 suggests the direction that he, DunnHumby and later Tesco would take in constructing lists [@OKeefe_1984]. The abstract for Humby's presentation pre-figures an ongoing trajectory for data mining techniques aimed at eliciting detailed information on individual customer preferences.

@@

While Tesco  succeeded in data-mining its customers,  and perhaps became the UK's biggest supermarket with the help of data-mining in the 1990s, the online shopping environment  in 2016 is a bit different. It is no longer organised around campaigns involving special offers or redemption of points for demographically segmented customer (DunnHumby made heavy use of UK Census data). It can no longer rely on placement of goods in carefully chosen locations in stores. It needs or at least might want to continuously up-sell and cross-sell to customers who might hardly ever be in the supermarket itself. If 'Tesco is the clear winner in the online grocery market, in fact it takes almost 50p of every £1 spent on food shopping on the internet' [How SEO helps Tesco to dominate the online grocery market](https://econsultancy.com/blog/64841-how-seo-helps-tesco-to-dominate-the-online-grocery-market)[@Silverwood-Cope_2014], then Tesco itself has undergone some kind of conversion?

@@

How could we characterise the shift from ClubCard data mining to online grocery markets?  Patel described her work at DunnHumby in terms of a shift in models from a 'rules-based list' to a 'relevancy model.' The relevancy model changed how the 'Have You Forgotten' list was constructed.   This shift is a typical of broader re-organisation of prediction in which existing predictive practices are being re-distributed and intensified in particular ways. The shift in models pivots on a more probabilistic structuring of lists.  The ostensible banality of the 'Have you forgotten?' recommendations betrays micrological signs of a topologically complex predictive infrastructure that subducts people and things in significant ways.

## Reconfiguring recommendations

HERE
